Recursive self-embedded vocal motifs in wild orangutans

Version 1

Abstract

Abstract

Recursive procedures that allow placing a vocal signal inside another of similar kind provide a neuro-computational blueprint for syntax and phonology in spoken language and human song. There are, however, no known vocal patterns among nonhuman primates arranged in self-embedded combinations that evince vocal recursion or potential insipient forms and neuro-procedures thereof, suggesting a neuro-cognitive transformation exclusive to humans. Here, we uncover that wild flanged male orangutan long calls show two hierarchical strata, wherein rhythmically isochronous call sequences are nested within self-similar isochronous call sequences. Remarkably, three unrelated recursive motifs occurred simultaneously in long calls, refuting that motifs resulted from three parallel linear procedures or that motifs were simple anatomical artifacts of bodily constrains. Findings represent a case of recursive hominid vocal production in the absence syntax, semantics, phonology or music. Second-order combinatorics, ‘sequences within sequences’, involving hierarchically organized and cyclically structured vocal sounds in ancient hominids may have preluded the evolution of recursion in modern language-able humans.

Introduction

Among the many definitions of recursion (Martins, 2012), the view that it represents the capacity to iterate a signal within a self-similar signal has crossed centuries and disciplines, from von Humboldt (1836) and Hockett (1960), to Mandelbrot (1980) and Chomsky (2010); from fractals in mathematics (Mandelbrot, 1980) to generative grammars in linguistics (Chomsky, 2010). Across varying terminologies, the common denominator across fields is that to re-curse (that is, to ‘re-invoke’) or re-iterate is an algorithmic “hack” to produce infinite signal states from a finite signal set. Although classically associated with syntax (Chomsky, 2010; Idsardi et al., 2018), recursive signal production and the resulting self-embedded structures have also been recognised in phonology (Bennett, 2018; Elfner, 2015; Kabak and Revithiadou, 2009; Nasukawa, 2015, 2020; Vogel, 2012) and (verbal and non-verbal) music (Jackendoff, 2009; Koelsch et al., 2013; Martins et al., 2017; Sharma and Chimalakonda, 2018). Given that language and music are uniquely human, this hints at a neuro-cognitive or neuro-computational transformation in the human brain that enabled the emergence of multiple open-ended communication systems in the human lineage, but seemingly, none other (Hauser et al., 2002). The absence of vocal structures in (nonhuman) primates that could help inform insipient or transitional recursive states, namely within the hominid family, has led some scholars to question altogether whether recursion was the result of natural selection (Berwick and Chomsky, 2019; Bolhuis and Wynne, 2009) and to rebut the role of shared ancestry and evolution as an incremental path-dependent process (cf. de Boer et al., 2020; Jackendoff and Pinker, 2005; Kershenbaum et al., 2014; Martins and Boeckx, 2019), favouring instead sudden “hopeful monster” mutant scenarios.

Decades-long debates on the evolution of recursion have ensued, carved around the successes and limitations of empirical comparative animal research. For example, perception and processing of syntax-like vocal combinatorics has been identified in some bird (Engesser et al., 2019, 2016; Gentner et al., 2006; Liao et al., 2022; Suzuki et al., 2016, 2017) and primate species (Jiang et al., 2018; Wang et al., 2015; Watson et al., 2020) but results’ interpretation has received various criticisms (Johan J Bolhuis et al., 2018; Bowling and Fitch, 2015; Corballis and Corballis, 2014; Rawski et al., 2021). For example, these animal studies have, thus far, almost exclusively focused on perception (cf. Ferrigno et al., 2020; Liao et al., 2022): animals recognizing or responding to recursive signal structures; however, processing operations of recursive signals are not necessarily recursive themselves, for example (Corballis and Corballis, 2014; Miyagawa, 2021). Nonetheless, critically, most research has focused on laboratory animals in artificial test settings and/or who were presented with synthetic stimuli after dedicated human training (Gentner et al., 2006; Jiang et al., 2018; Liao et al., 2022; Watson et al., 2020), including the few exceptions based on production instead of perception (Ferrigno et al., 2020; Liao et al., 2022). This opens comparative research to two vital drawbacks. First, results are mute about evolutionary precursors and processes. Settings and stimuli have not been those to which species adapted over evolutionary time, with most species tested thus far being too distantly related to humans to allow inferences about recursion emergence among hominids in the first place. That animals may be able to recognise recursive structures or engage in non-vocal recursive action after undergoing fixed training protocols with human-designed stimuli will never per se indicate whether recursion has been selected in the wild, including human ancestors.

Second, results have been slanted by the dominant classic theory that, counter-intuitively, disfavours a gradual evolutionary scenario for recursion and instead conceives it as a punctuated event (Berwick and Chomsky, 2019; Bolhuis et al., 2015; Bolhuis and Wynne, 2009). Namely, experimental stimuli have consisted of artificial recursive signal sequences organized along a single temporal scale (though not structurally linear), similarly with how Merge and syntax operate. However, recursive signal structures can also unfold in other manners, such as across nested temporal scales and in the absence of semantics (Fitch, 2017a), as in music. But these remain utterly unprobed and untested.

An approach to the production of recursive signals, complementary to perceptual studies, is thus, desirable. Data from wild primates in particular may help infer signal patterns that were recursive in some degree or kind in an extinct past and subsequently moulded into the recursive structures observed today in modern humans. By virtue of their own primitive nature, proto-recursive structures did not likely fall within modern-day classifications, and thus, will often fail to be predicted based on assumptions guided by full-fledged language (Kershenbaum et al., 2014; Miyagawa, 2021). To this end, a structural approach is particularly advantageous. First, no prior assumptions are required about species’ cognitive capacities. These are directly inferred from how signal sequences are organised. For example, Chomsky’s definition of recursion (Chomsky, 2010) can generate non-self-embedded signal structures, but these would be operationally undetectable amongst other signal combinations. Second, no prior assumptions are required about signal meaning. There are no certain parallels with semantic content and word meaning in animals, but analyses of signal patterning allows to identify similarities between non-semantic (nonhuman) and semantic (human) combinatoric systems (Lipkind et al., 2013; Sainburg et al., 2019). Therefore, the search for recursion can be made in the absence of meaning-base operations, such as Merge, and more generally, semantics and syntax. Third, no prior assumptions are required about signal function. Under punctuated or gradual evolutionary hypotheses, ancestral signal function (whether cooperative, competitive or otherwise) is expected to have derived or been leveraged by its proto-recursive structure. Otherwise, once present and despite its origin process, recursion would not have been fixated among human ancestral populations. A structural approach opens, therefore, the field to untapped signal diversity in nature and yet unrecognised bona fide combinatoric states within the human clade.

Here, undertaking an explorative structural approach to recursion, we provide evidence for recursive self-embedded vocal patterning in a (nonhuman) great ape, namely, in the long calls of flanged orangutan males in the wild. We conducted precise rhythm analyses (De Gregorio et al., 2021; Roeske et al., 2020) of 66 long call audio recordings produced by 10 orangutans (Pongo pygmaeus wurmbii) across approximately 2510 observation hours at Tuanan, Central Kalimantan, Indonesian Borneo. We identified 5 different element types that comprise the structural building blocks of long calls in the wild (Hardus et al., 2009; Lameira and Wich, 2008), of which the primary type are full pulses (Fig. 1A). Full pulses do not, however, always exhibit uninterrupted vocal production throughout a long call [as during a long call’s climax (Spillmann et al., 2010)] but can break-up into 4 different “sub-pulse” element types: (i) grumble sub-pulses [quick succession of staccato calls that typically constitute the first build-up pulses of long calls (Hardus et al., 2009)], (ii) sub-pulse transitory elements and (iii) pulse bodies (typically constituting pulses before and/or after climax pulses) and (iv) bubble sub-pulses (quick succession of staccato calls that typically constitute the last tail-off pulses of long calls) (Fig. 1A). We characterised long calls’ full- and sub-pulses’ rhythmicity to determine if orangutan long calls present a re-iterated structure across different hierarchical strata. We extracted inter-onset-intervals (IOIs; i.e. time difference between the start of a vocal element and the preceding one - tk) from 8993 vocal long call elements (Fig. 1A): 1930 full pulses (1916 after filtering for 0.025<tk<5s), 757 grumble sub-pulses (731), 1068 sub-pulse transitory elements (374), 816 pulse bodies (11) and 4422 bubble sub-pulses (4193). From the extracted IOIs, we calculated their rhythmic ratio by dividing each IOI by its duration plus the duration of the following interval. We then computed the distribution of these ratios to ascertain whether the rhythm of long call full and sub-pulses presented natural categories, following published protocols (De Gregorio et al., 2021; Roeske et al., 2020) (Fig. 1B, C, D).

Organization and rhythmic features of orangutans’ long calls.

(A) On top, the spectrogram of a Full Pulse and its organization in Sub-Pulses (e.g., Grumble sub-pulses). Below are the spectrograms of the three other sub-element types: Sub-pulse transitory elements, Pulse bodies and Bubble sub-pulses. Bars on the top of each spectrogram schematically quantify durations of inter-onset intervals (tk): dark green denotes the higher-level of organization (Full pulse). Orange (in the inset) and light green (bottom right) denote the lower-level organization (sub-pulse element types).

(B) Probability density function showing the distributions of the inter-onset-intervals (tk) for each of the long call element types.

(C) The distributions on the left show rhythm ratios (rk) per element type as calculated on 12 flanged males for a total of 1915 Full-pulses and 5309 sub-pulses. Solid sections of the curves indicate on-isochrony rk values; striped sections indicate off-isochrony rk values. A solid white line indicates the 0.5 rk value corresponding to isochrony. White dotted lines denote the on-isochrony peak value extracted from the probability density function. On the right, a bar plot per each element type shows the percentage of observations (rk) falling into the on-isochrony boundaries (solid bars) or on off-isochrony boundaries (striped bars). The number of on-isochrony rk is significantly larger (GLMM, Full vs Null: Chisq=2717.543, p<0.001) than the number of off-isochrony rk for all long call element types (Full pulse: t-ratio=-25.164, p<0.001; Bubble sub-pulse: t-ratio= -30.694, p<0.001; Grumble sub-pulse: t-ratio=-14.526, p<0.001; Sub-pulse transitory element: t-ratio=-3.148, p<0.001). Pulse body showed no rk values falling within the on-off-isochrony boundaries.

(D) Distribution of a variable calculated as the ratio between the tk of a sub-pulse and the tk of the corresponding higher level of organization, the Full Pulse. We report the peak value of the curve (0.046) and tested the significance of the extent of the central quartiles, which was significantly smaller than peripheral quartiles (Wilcoxon signed-rank test: W=2272, p<0.001).

Results

The density probability function of orangutan full pulses showed one peak (rk=0.493) in close vicinity to a theoretically pure isochronic rhythm, that is, full pulses were regularly paced at 1:1 ratio, following a constant tempo along the long call (Fig. 1C). Our model (GLMM, full model vs null model: Chisq=298.2876, df=7, p<0.001; see Supplementary Materials) showed that pulse type, range of the curve (on-off-isochrony), and their interaction, had a significant effect on the count of rk values. In particular, full pulses’ isochronous peak tested significant (t.ratio=-15.957, p<0.0001), that is, the number of rk values falling inside on-isochrony range was significantly higher than the number of rks falling inside the off-isochrony range (Fig. 1C). Critically, three (of the four) orangutan sub-pulse element types – grumble sub-pulses, sub-pulse transitory elements and bubble sub-pulses – also showed significant peaks (grumble sub-pulses: t.ratio = -5.940, p<.0001; sub-pulse transitory elements: t.ratio=-4.048, p=0.0001; bubble sub-pulses: t.ratio= - 10.640, p<.0001) around pure isochrony (peak rk: grumble sub-pulses = 0.501; sub-pulse transitory elements=0.495; bubble sub-pulses=0.502; Fig. 1C). That is, sub-pulses were regularly paced within regularly paced full pulses, denoting isochrony within isochrony (Fig. 1C) at different average tempi (mean tk (sd): full pulses=1.696 (0.508); grumble sub-pulses=0.118 (0.111); sub-pulse transitory elements=0.239 (0.468); bubble sub-pulses= 0.186 (0.292); Fig. 1B). Overall, sub-pulses’ tk was equivalent to 0.046 of their comprising full-pulses (Fig. 1D), which put sub-pulses at an approximate ratio of 1:20 relative to that of full-pulses, the smallest categorical temporal rhythmic interval registered thus far in a vertebrate (De Gregorio et al., 2021; Roeske et al., 2020). Permuted discriminant function analyses (Mundry and Sommer, 2007) (crossed, in order to control for individual variation) in R (Team, 2013) based on seven acoustic measures extracted from grumble, transitory and bubble sub-pulses confirmed that these represented indeed distinct sub-pulse categories, where the percentage of correctly classified selected cases (62.7%) was significantly higher (p=0.001) than expected (37%).

Discussion

Orangutan long call nested rhythmic patterns reveal self-similar embedded isochrony in the vocal production of a wild great ape, notably, with two discernible structural strata – the full- and sub-pulse level – and three non-exclusive rhythmic arrangements in the form of [isochronyA [isochronya,b,c]].

Human and nonhuman great apes have similar auditory capacities (Quam et al., 2015). There are no identified skeletal differences in inner ear anatomy that could suggest significantly distinct sound sensitivity, resolution or activation thresholds in the time domain (Quam et al., 2015). Humans perceive an acoustic pulse as a continuous pitch, instead of a rhythm, at rates higher than 30 Hz (i.e., 30 beats per second). Long call sub-pulses exhibited average rhythms at ∼9.263 (3.994) Hz [i.e., tk=0.184 (0.303)s]. Therefore, hominid ear anatomy offers strong confidence that orangutans, like humans (and other great apes) perceive sub-pulse rhythmic motifs as such, i.e., as a train of signals, instead of one uninterrupted signal. Assuming otherwise would imply that auditory time-resolution differ by more than one order of magnitude between humans and other great apes in the absence of obvious anatomical culprits.

The simultaneous occurrence of non-exclusive recursive patterns excludes the likelihood that orangutans concatenate long calls and their subunits in linear structure without any recursive processes. To generate the observed vocal motifs as linear structure, three independent neuro-computational procedures would need to run in parallel to generate distinct isochronic rhythms at the sub-pulse level, whilst being indistinguishable, transposable and/or interchangeable at the pulse level without interference, which would be unlikely, if theoretically possible at all. Non-exclusive simultaneous recursive patterns also help exclude the probability that recursion was the primary by-product of anatomic constrains, such as breath length, heart beat and other physiological rhythmic processes or movements (Pouw et al., 2020). Such processes could generate various simultaneous rhythmic patterns; however, these would be expected to be in the form of harmonics of the same base “carrier” rhythm. Yet, the three observed rhythmic arrangements at the sub-pulse level were not related to the pulse level by any small integer ratios. Together, this strongly suggests that the observed recursive self-embedded motifs are most likely generated by a recursive neurological procedure or recursion algorithm.

Recursive self-embedded vocal motifs in orangutans indicate that vocal recursion among hominids is not exclusive to human vocal and cognitive systems. This is not to suggest that they exhibit all properties that recursion exhibits in modern language-able humans. Such expectation would be unreasonable, as it would imply that no evolution occurred in >10 million years since the split between the orangutan and human phylogenetic lineages. Thus, any apparent differences with recursion in today’s syntax, phonology, or music do not invalidate the probability that the reported vocal motifs represent an ancient, or potential ancestral, state for the ensuing evolution of recursion along the human clade.

Recursion and fractal phenomena are prevalent across the universe. From celestial and planetary movement to the splitting of tree branches and river deltas, and the morphology of bacteria colonies; Patterns within self-similar patterns are the norm, not the exception. This makes the seeming singularity of human recursion amongst vertebrate signals only the more enigmatic. Our findings indicate that ancient vocal patterns organized across nested structural strata were likely present in ancestral hominids. Recursive vocal production likely predated the evolutionary emergence of (spoken) language within the Hominid family and along the human lineage. This data-driven possibility poses new evolutionary trajectories and timelines distinct from the classic theoretical notions that the advent of recursion was a saltational all-or-nothing event that took place only recently in modern humans (Berwick and Chomsky, 2019). Gradual cumulative evolutionary scenarios for the emergence and evolution of recursion (de Boer et al., 2020; Martins and Boeckx, 2019) can and should be considered and investigated, with the advantage that living primate models in general, and natural hominid behaviour in particular, offer superior empirical and comparative validity beyond purely theoretical considerations (Lameira et al., 2021). A combination of both approaches will likely pay the highest heuristic dividends. For example, future research may implement playback experiments designed to clarify whether or how great apes cognitively represent natural recursive self-embedded motifs as informed by previous designs and theoretical considerations (Corballis and Corballis, 2014; Engesser et al., 2016; Watson et al., 2020).

Implications for the evolution of recursion and cognition

The presence of recursive self-embedded vocal motifs in orangutans carries four major implications for the evolution of recursion and cognition. First, much ink has been laid on the topic, yet, the possibility of self-embedded isochrony, or alternatively, non-exclusive self-embedded patterns occurring within the same signal sequence, has on no account been formulated or conjectured as a possible state of recursive signalling, be it in vertebrates, mammals, primates or otherwise, extant or extinct. This suggests that controversy may have thus far been underscored by data-poor circumstances, namely, lack of rigorous and detailed knowledge about call combinations in wild primates in general, and great apes in particular. The presence of recursive vocal patterns in a wild great ape in the absence of syntax, phonology or music helps prevent prior assumptions that recursion precursors ought, or should be expected to, operate as modern-day recursion. Orangutan recursive vocal patterns provide the first data point and open a new charter for possible insipient or transitional states within the hominid family that, whilst inherently distinct from modern-day recursion, were nonetheless homologous to it and fully functional in their own right. The open discussion about what kind of properties would or could make a structure proto-recursive or not will be essential to move the state-of-knowledge past antithetical, jointly exhaustive or mutually exclusive options of what recursion is, how it emerged and evolved.

Second, primate loud calls are functionally analogous to, for instance, bird and whale song. Accordingly, sophisticated signalling in far-related species does not undermine the value of great apes as living models for the study of language and its underpinning neuro-motoric operations, as often claimed (Berwick and Chomsky, 2019; Johan J. Bolhuis et al., 2018; Bolhuis and Wynne, 2009; Lattenkamp and Vernes, 2018; Suzuki et al., 2018; Vernes et al., 2021) based on the false assumption that great ape vocal research has been at a standstill for the last half-century (cf. Belyk and Brown, 2017; Bianchi et al., 2016; Crockford et al., 2004; Hopkins et al., 2007; Lameira, 2017; Lameira et al., 2016, 2015, 2013; Lameira and Shumaker, 2019; Pereira et al., 2020; Russell et al., 2013; Staes et al., 2017; Taglialatela et al., 2012; Watson et al., 2015; Wich et al., 2009, 2012). Our findings invite results across lineages to be tested with primates in order to explicitly assess their evolutionary relevance within the human clade. This will help avoid prematurely interpreting absence of evidence for evidence of absence in primates.

Third, our findings suggest that, despite criticism, recursive perceptual capacities identified in primates (Watson et al., 2020) may be factual and likely evolved to subserve vocal signals in the species’ natural repertoire (though potentially yet unrecognized). This invites for a renewed interest and re-analysis of primate signalling behaviour in the wild (Gabrić, 2021). However, findings also show that it may be too hasty to discuss whether perceptual capacities in primates or birds are equivalent to those engaged in syntax (Watson et al., 2020) or phonology (Rawski et al., 2021). Such classifications may be putting the proverbial cart before the horse; they are based on untested assumptions (e.g., that syntax and phonology evolved as separate “modules”, that one attained modern form before the other, et cetera) that may not have applied to proto-recursive ancestors (Kershenbaum et al., 2014; Miyagawa, 2021).

Forth, given that isochrony universally governs music and that recursion is a feature of music, findings suggest a possible evolutionary link between great ape loud calls and vocal music. Loud calling is an archetypal trait in primates (Wich and Nunn, 2002) but is absent in modern humans. Our findings suggest this may not be coincidental. Great ape loud calling may have preceded and subsequently transmuted into modern recursive vocal structures in humans. Given their conspicuousness, loud calls represent one of the most studied aspects of primate vocal behaviour (Wich and Nunn, 2002), but their rhythmic patterns have seldom been characterized with precision (Clink et al., 2020; De Gregorio et al., 2021; Gamba et al., 2016). Besides our analyses, there are remarkably few confirmed cases of isochrony in great apes (but see Raimondi et al., 2023), but the behaviours that have been rhythmically measured with accuracy have been implicated in the evolution of percussion (Fuhrmann et al., 2015) and musical expression (Dufour et al., 2015; Hattori and Tomonaga, 2020), such as social entrainment in chimpanzees in connection with the origin of dance (Lameira et al., 2019) [a capacity once also assumed to be neurologically impossible in great apes (Fitch, 2017b; Patel, 2014)]. This opens the possibility that recursive vocal production and supporting neural procedures were first and foremost a feature of proto-musical expression in human ancestors, later recruited and “re-engineered” for the generation of linguistic combinatorics.

Future studies outlining the distribution of isochrony across primate (vocal) behaviour offer promising new paths to empirically assess the evolution of recursive signal structures in music and language and will help move the needle forward on one the most tantalizing riddles in the evolution of language and cognition. These crucial data and insights will materialise if, as stewards of our planetary co-habitants, humankind secures the survival of these species and the preservation of their natural wild habitat (Estrada et al., 2022, 2017; Laurance, 2013; Laurance et al., 2012).

Methods and Materials

Study site

We conducted our research at the Tuanan Research Station (2°09′S; 114°26′E), Central Kalimantan, Indonesia. Long calls were opportunistically recorded from identified flanged males (Pongo pygmaeus wurmbii) using a Marantz Analogue Recorder PMD222 in combination with a Sennheiser Microphone ME 64 or a Sony Digital Recorder TCD-D100 in combination with a Sony Microphone ECM-M907.

Acoustic data extraction

Audio recordings were transferred to a computer with a sampling rate of 44.1 kHz. Seven acoustic measures were extracted directly from the spectrogram window (window type: Hann; 3 dB filter bandwidth: 124 Hz; grid frequency resolution: 2.69 Hz; grid time resolution: 256 samples) by manually drawing a selection encompassing the complete long call (sub)pulse from onset to offset, using Raven interactive sound analysis software (version 1.5, Cornell Lab of Ornithology). These parameters were duration(s), peak frequency (Hz), peak time, peak frequency contour average slope (Hz), peak frequency contour maximum slope (Hz), average entropy (Hz), signal-to-noise ratio (NIST quick method). Please see software’s documentation for full description of parameters (https://ravensoundsoftware.com/knowledge-base/pitch-tracking-frequency-contour-measurements/). Acoustic data extraction complemented the classification of long calls elements, both at the pulse and sub-pulse levels, based on close visual and auditory inspection of spectrograms, both based on elements’ distinctiveness between each other as well as in relation to the remaining catalogued orangutan call repertoire (Hardus et al., 2009) (see also supplementary audio files). Of these parameters, duration and peak frequency in particular have been shown to be resilient across recording settings(Lameira et al., 2013) and to adequately represent variation in the time and frequency axes (Lameira et al., 2017).

Rhythm data analyses

Inter-onset-intervals (IOI’s = tk) were only calculated from the begin time (s) of each full-and sub-pulse long call elements using Raven interactive sound analysis software, as above explained. tk was calculated only from subsequent (full/sub) pulse elements of the same type. Ratio values (rk) were calculated as tk/(tk+tk+1). Following the methodology of Roeske et al., 2020 and De Gregorio et al. 2021, to assess the significance of the peaks around isochrony (corresponding to the 0.5.rk value), we counted the number of rks falling inside on-isochrony ranges (0.440 < rk < 0.555) and off-isochrony ranges (0.400 < rk < 0.440 and 0.555 < rk < 0.600), symmetrically falling at the right and left sides of 1:1 ratios (0.5 rk value). We tested the count of on-isochrony rks versus the count of off-isochrony rks, per pulse type, with a GLMM for negative-binomial family distributions, using glmmTMB R library. In particular, we built a full model with the count of rk values as the response variable, the pulse type in interaction with the range the observation fell in (on- or off-isochrony) as predictors. We added an offset weighting the rk count based on the width of the bin. The individual contribution was set as random factor. We built a null model comprising only the offset and the random intercepts. We checked the number of residuals of the full and null models, and compared the two models with a likelihood ratio test (Anova with “Chisq” argument). We calculated p-values for each predictor using the R summary function and performed pairwise comparisons for each level of the explanatory variables with emmeans R package, adjusting all p-values with Bonferroni correction. We checked normality, homogeneity (via function provided by R. Mundry), and number of the residuals. We checked for overdispersion with performance R package (Lüdecke et al., 2020). Graphic visualization was prepared using R (Team, 2013) packages ggplot2 (Wickham, 2009) and ggridges (Wilke, 2022). Data reshape and organization were managed with dplyr and tidyr R packages.

Acoustic data analyses

Permutated discriminant function analysis with cross classification was performed using R and a function provided by Roger Mundry (Mundry and Sommer, 2007). The script was: pdfa.res=pDFA.crossed (test.fac=“Sub-pulse-type”, contr.fac=“Individual.ID”, variables=c(“Delta.Time”, “Peak.Freq”, “Peak.Time”, “PFC.Avg.Slope”, “PFC.Max.Slope”, “Avg.Entropy”, “SNR.NIST.Quick”), n.to.sel=NULL, n.sel=100, n.perm=1000, pdfa.data=xdata). These analyses assured that long call elements, at the pulse and sub-pulse level, indeed represented biologically distinct categories.

eLife Assessment

The paper provides a useful and novel application of recursion theory to the long call vocalisations of orangutans to demonstrate repetitive, rhythmic sub-structuring. However the evidence provided to support the major claims of the paper is currently incomplete. Specifically, it is not yet clear how the rhythmic structuring found in these long calls is more similar to human language recursion per se rather than isochrony as a broader, more common phenomenon.

Reviewer #1 (Public Review):

This study investigates the structuring of long calls in orangutans. The authors demonstrate long calls are structured around full pulses, repeated following a regular tempo (isochronic rhythm). These full pulses are themselves structured around different sub-pulses, themselves repeated following an isochronic rhythm. The authors argue this patterning is evidence for self-embedded, recursive structuring in orangutang long calls.

The analyses conducted are robust and compelling and they support the rhythmicity the authors argue is present in the long calls. Furthermore, the authors went above and beyond and confirmed acoustically the sub-categories identified were accurate.

However, I believe the manuscript would benefit from a formal analysis of the specific recursive patterning occurring in the long call. Indeed, as of now, it is difficult for the reader to identify what the authors argue to be recursion and distinguish it from simple repetitions of motifs, which is essential. Although the authors already discuss briefly why linear patterning is unlikely, the reader would benefit from expanding on this discussion section and clarifying the argument here (a lay terminology might help). I believe an illustration here might help.

In the same logic, I believe a tree similar to the trees used in linguistics to illustrate hierarchical structuring would help the reader understand the recursive patterning in place here. This would also help get the "big picture", as Fig 1A is depicting a frustratingly small portion of the long call.

Notwithstanding these comments, this paper would provide crucial evidence for recursion in the vocal *production* of a non-human ape species. The implication it would have would represent a key shift in the field of language evolution. The study is very elegant and well-constructed. The paper is extremely well written, and the point of view adopted is original, well-argued and compelling.

Reviewer #2 (Public Review):

I am not qualified to judge the narrow claim that certain units of the long calls are isochronous at various levels of the pulse hierarchy. I will assume that the modelling was done properly. I can however say that the broad claims that (i) this constitutes evidence for recursion in non-human primates, (ii) this sheds light on the evolution of recursion and/or language in humans are, when not made trivially true by a semantic shift, unsupported by the narrow claims. In addition, this paper contains errors in the interpretation of previous literature.

The main difficulty when making claims about recursion is to understand precisely what is meant by "recursion" (arguably a broader problem with the literature that the authors engage with). The authors offer some characterization of the concept which is vague enough that it can include anything from "celestial and planetary movement to the splitting of tree branches and river deltas, and the morphology of bacteria colonies". With this appropriately broad understanding, the authors are able to show "recursion" in orangutans' long calls. But they are, in fact, able to find it everywhere. The sound of a plucked guitar string, which is a sum of self-similar periodic patterns, count as recursive under their definition as well.

One can only pick one's definition of recursion, within the context of the question of interest: evolution of language in humans. One must try to name a property which is somewhat specific to human language, and not a ubiquitous feature of the universe we live in, like self-similarity. Only after having carved out a sufficiently distinctive feature of human language, can we start the work of trying to find it in a related species and tracing its evolutionary history. When linguists speak of recursion, they speak of in principle unbounded nested structure (as in e.g. "the doctor's mother's mother's mother's mother ..."). The author seems to acknowledge this in the first line of the introduction: "the capacity to *iterate* a signal within a self-similar signal" (emphasis added). In formal language theory, which provides a formal and precise definition of one notion of recursivity appropriate for human language, unbounded iteration makes a critical difference: bounded "nested structures" are regular (can be parsed and generated using finite-state machines), unbounded ones are (often) context-free (require more sophisticated automaton). The hierarchy of pulses and sub-pulses only has a fixed amount of layers, moreover the same in all productions; it does not "iterate".

Another point is that the authors don't show that the constraints that govern the shape of orangutans long calls are due to cognitive processes. Any oscillating system will, by definition, exhibit isochrony. For instance, human trills produce isochronouns or near isochronous pulses. No cognitive process is needed to explain this; this is merely the physics of the articulators. Do we know that the rhythm of the pulses and sub-pulses in orangutans is dictated by cognition as opposed to the physics of the articulators?

Even granting the authors' unjustified conclusion that wild orangutans have "recursive" structures and that these are the result of cognition, the conclusions drawn by the authors are too often fantastic leaps of induction. Here is a cherry-picked list of some of the far-fetched conclusions:

- "our findings indicate that ancient vocal patterns organized across nested structural strata were likely present in ancestral hominids". Does finding "vocal patterns organized across nested structural strata" in wild orangutans suggest that the same were present in ancestral hominids?

- "given that isochrony universally governs music and that recursion is a feature of music, findings (sic.) suggest a possible evolutionary link between great ape loud calls and vocal music". Isochrony is also a feature of the noise produced by cicadas. Does this suggest an evolutionary link between vocal music and the noise of cicadas?

Finally, some passages also reveal quite glaring misunderstandings of the cited literature. For instance:

- "Therefore, the search for recursion can be made in the absence of meaning-base operations, such as Merge, and more generally, semantics and syntax". It is precisely Chomsky's (disputable) opinion that the main operation that govern syntax, Merge, has nothing to do with semantics. The latter is dealt within a putative conceptual-intentional performance system (in Chomsky's terminology), which is governed by different operations.

- "Namely, experimental stimuli have consisted of artificial recursive signal sequences organized along a single temporal scale (though not structurally linear), similarly with how Merge and syntax operate". The minimalist view advocated by Chomsky assumes that mapping a hierarchichal structure to a linear order (a process called linearizarion) is part of the articulatory-perceptual system. This system is likewise not governed by Merge and is not part of "syntax" as conceived by the Chomskyan minimalists.

Author Response

The following is the authors’ response to the original reviews.

Thank you very much for forwarding these two important reviews on our paper. Please find hereby our point-by-point responses addressing the ideas, arguments and points of concern raised by the reviewers. We provide explanation of how these points have been incorporated in the paper.

We feel the review process has been a useful exercise and that the paper has greatly benefited in terms of clarity and accessibility. It is our hope that our findings may ignite renewed interest on unexplored and “unexpected” aspects of great ape vocal communication, inspire novel research, and invite bold new advances on the long-standing puzzle of language origins and evolution. In several relevant sections, we have also sought to explicitly address the point of doubt raised in eLife’s editorial assessment, published alongside the reviewed preprint of our paper. The editorial assessment stated that “…However the evidence provided to support the major claims of the paper is currently incomplete. Specifically, it is not yet clear how the rhythmic structuring found in these long calls is more similar to human language recursion per se rather than isochrony as a broader, more common phenomenon.” To directly clarify this point, we provide now various examples of how recursion is distinct from repetition, using everyday objects for an intuitive understanding (e.g., lines 43-51). We have also expanded the discussion to better contextualise and clarify the implications of our findings on language evolution theory. We hope this will help addressing the implicit request for clarification in the previous editorial assessment.

Thank you very much for your kind and dedicated attention in the processing of our study.

Public Reviews:

Reviewer #1 (Public Review):

This study investigates the structuring of long calls in orangutans. The authors demonstrate long calls are structured around full pulses, repeated following a regular tempo (isochronic rhythm). These full pulses are themselves structured around different sub-pulses, themselves repeated following an isochronic rhythm. The authors argue this patterning is evidence for self-embedded, recursive structuring in orangutang long calls.

The analyses conducted are robust and compelling and they support the rhythmicity the authors argue is present in the long calls. Furthermore, the authors went above and beyond and confirmed acoustically the sub-categories identified were accurate.

We thank the reviewer for this important support regarding our methods and findings.

However, I believe the manuscript would benefit from a formal analysis of the specific recursive patterning occurring in the long call. Indeed, as of now, it is difficult for the reader to identify what the authors argue to be recursion and distinguish it from simple repetitions of motifs, which is essential.

We agree with the reviewer that the distinction between repetition and recursion is very important for the adequate interpretation of our findings. Following the reviewer’s point (and the Editorial Assessement), we have now rephrased several passages in the initial paragraph of the paper for added clarity, where recursion is introduced and explained. We now also provide various new examples of recursion in everyday life and popular culture to better illustrate in an easy and accessible way the fundamental nature of recursion. We then use two of these common examples (computer folders and Russian dolls) to specifically distinguish repetition from recursion.

Although the authors already discuss briefly why linear patterning is unlikely, the reader would benefit from expanding on this discussion section and clarifying the argument here (a lay terminology might help).

Corrected accordingly.

I believe an illustration here might help. In the same logic, I believe a tree similar to the trees used in linguistics to illustrate hierarchical structuring would help the reader understand the recursive patterning in place here. This would also help get the "big picture", as Fig 1A is depicting a frustratingly small portion of the long call.

We completely understand the reviewer’s concern here. As proposed by the reviewer, and in addition to changes in the Introduction (see above) and Discussion (see below), we have now added a new figure in the Discussion to help the reader get the “big picture” of our findings.

We have also made revisions throughout the Introduction and Discussion to simplify the text, clarify our exposition and facilitate the reader better and intuitively understand the nature and relevance of our results.

Notwithstanding these comments, this paper would provide crucial evidence for recursion in the vocal production of a non-human ape species. The implication it would have would represent a key shift in the field of language evolution. The study is very elegant and well-constructed. The paper is extremely well written, and the point of view adopted is original, well-argued and compelling.

We are humbled by the reviewer’s words, and we thank the reviewer for attributing these qualities to our paper. This feedback reassures us of the disruptive potential that these and similar future findings may have on our understanding of language evolution.

Reviewer #2 (Public Review):

I am not qualified to judge the narrow claim that certain units of the long calls are isochronous at various levels of the pulse hierarchy. I will assume that the modelling was done properly. I can however say that the broad claims that (i) this constitutes evidence for recursion in non-human primates, (ii) this sheds light on the evolution of recursion and/or language in humans are, when not made trivially true by a semantic shift, unsupported by the narrow claims. In addition, this paper contains errors in the interpretation of previous literature.

We report the first confirmed case of “vocal sequences within vocal sequences” in a wild nonhuman primate, namely a great ape. The currently prevailing models of language evolution often rest on the (purely theorical) premise that such structures do not exist in any animal bar humans. We find the discovery of such structures in a wild great ape exciting, remarkable, and promising. We regret that the reviewer does not share this sentiment with us. We feel that the statement that these findings are trivial and narrow is unfounded.

In order to clarify and better communicate the significance of our findings, we now explain in more detail in the Introduction and Discussion how the discovery of nested isochrony in wild orangutans promises to stimulate new series of studies in nature and captivity. Our findings dovetail nicely with previous captive studies that have shown that animals can learn how to recognise recursive patterns and invite new research efforts for the investigation of recursive abilities in the wild and in the absence of human priming and in nonhuman primates.

The main difficulty when making claims about recursion is to understand precisely what is meant by "recursion" (arguably a broader problem with the literature that the authors engage with). The authors offer some characterization of the concept which is vague enough that it can include anything from "celestial and planetary movement to the splitting of tree branches and river deltas, and the morphology of bacteria colonies". With this appropriately broad understanding, the authors are able to show "recursion" in orangutans' long calls. But they are, in fact, able to find it everywhere.

The reviewer is correct in highlighting that recursion is ubiquitous in nature and this is something that we explicitly state in the paper. This only makes it the more surprising that, when it comes to vocal combinatorics, recursion has only been described in human language and music, but in no other animals. If studies providing such evidence are known to reviewer, we kindly request their corresponding references.

In the new revised version, we have paid attention to this aspect raised by the reviewer, and we have sought to disambiguate that our observations pertain to temporal recursion. This clarification will hopefully allow a better understanding of our results.

The sound of a plucked guitar string, which is a sum of self-similar periodic patterns, count as recursive under their definition as well.

The example pointed out here by reviewer is factually correct; sound harmonics represent a recursive pattern of a fundamental frequency. (In fact, we explain this phenomenon in the Discussion.) The reviewer’s comment seems to offer an analogy to oscillatory phenomena in the physiology of the vocal folds, and so, it is misplaced with regards to our present study, which focused vocal sequences. Admittedly, this misinterpretation may have been implicitly caused by our wording and we apologise for this. We now refer to “vocal combinatorics” instead of “vocal production” throughout the paper to avoid the reader considering that our findings pertain to the physiology of the vocal folds.

One can only pick one's definition of recursion, within the context of the question of interest: evolution of language in humans. One must try to name a property which is somewhat specific to human language, and not a ubiquitous feature of the universe we live in, like self-similarity. Only after having carved out a sufficiently distinctive feature of human language, can we start the work of trying to find it in a related species and tracing its evolutionary history. When linguists speak of recursion, they speak of in principle unbounded nested structure (as in e.g., "the doctor's mother's mother's mother's mother ..."). The author seems to acknowledge this in the first line of the introduction: "the capacity to iterate a signal within a self-similar signal" (emphasis added). In formal language theory, which provides a formal and precise definition of one notion of recursivity appropriate for human language, unbounded iteration makes a critical difference: bounded "nested structures" are regular (can be parsed and generated using finite-state machines), unbounded ones are (often) context-free (require more sophisticated automaton). The hierarchy of pulses and sub-pulses only has a fixed amount of layers, moreover the same in all productions; it does not "iterate".

The reviewer explains here how recursion, in its fully fledged form in modern language(s), is defined by linguistics. We fully agree and do not contest such descriptions and definitions in any way. These descriptions and definitions aim to describe how recursion operates today, not how it evolved. Nor do these descriptions and definitions generate data-driven, testable predictions about precursors or proto-states of recursion as used by modern language-able humans. This is scientifically problematic and heuristically unsatisfying regarding the open question of language evolution.

Following human-specific definitions for recursion, as proposed by the reviewer, cannot per se be used to undertake a comparative approach to evolution because they leave nothing to compare recursion with in other (wild) species. Using human-specific definitions unavoidably leads to black-and-white notions that language is always absolutely present in humans and always absolutely absent in other animals, regardless of their degree of relatedness to humans. It is unpreventable that these descriptions flout foundational principles of evolution, such as descent with modification and shared ancestry.

This conceptual problem is not new. Less than a century ago, it was believed that humans were the only tool-user (thousands of examples are known today in nonhuman animals, including fish and invertebrates), and later, that humans were the only cultural animal (today it is known that migrating caribou and fruit flies can establish traditions based on social learning). We must follow in the footsteps of those who have helped redefine human nature in the past. As famously stated by Louis Leakey when presented with evidence for chimpanzee tool-use collected by Jane Goodall, “Now we must redefine tool, redefine man, or accept chimpanzees as human”. Therefore, as a matter of course, we must redefine recursion, embracing empirically (other than purely theoretically) definitions that allow recursion to take on forms and functions different from that of modern language-able humans.

Another point is that the authors don't show that the constraints that govern the shape of orangutans long calls are due to cognitive processes.

The reviewer is indeed correct. This does not, however, refute our findings. We do not directly show that cognitive processes govern recursion in orangutan long calls. Instead, we show that the observed patterns cannot be explained by simple bodily or motoric processes, excluding therefore low-level explanations. With more than 50 years of accumulated field experience in primatology, this was the only possible way that our team found to go about conducting research and analyses on natural behaviour, in the wild, with a critically endangered primate. We would be very interested in learning from the reviewer what ethical and non-invasive methods, specific locations in the wild, and type of behavioural or socio-ecological data could be otherwise viably used to demonstrate what the reviewer requests. If other scientists believe that the patterns observed in wild orangutan long calls – three independent, but simultaneously-occurring recursive motifs – can be generated based on low-level physiological mechanisms alone, the burden of proof resides with them.

Any oscillating system will, by definition, exhibit isochrony.

We disagree with this statement. The example provided above by the reviewer him/her-self disproves the statement: a guitar string when struck is an oscillating system but it is not isochronic nor is it combinatorial. Isochrony cannot be established with single events, only with event sequences (in practice, ideally >3).

For instance, human trills produce isochronouns or near isochronous pulses. No cognitive process is needed to explain this; this is merely the physics of the articulators. Do we know that the rhythm of the pulses and sub-pulses in orangutans is dictated by cognition as opposed to the physics of the articulators?

The reviewer seems to misinterpret our results here. Our focus is on vocal combinatorics, not vocal fold oscillation (see previous response). We have now reworded all instances where the text could be unclear.

Even granting the authors' unjustified conclusion that wild orangutans have "recursive" structures and that these are the result of cognition, the conclusions drawn by the authors are too often fantastic leaps of induction. Here is a cherry-picked list of some of the far-fetched conclusions:

"our findings indicate that ancient vocal patterns organized across nested structural strata were likely present in ancestral hominids". Does finding "vocal patterns organized across nested structural strata" in wild orangutans suggest that the same were present in ancestral hominids?

Following the reviewer’s comment, we have now rephrased and toned down this passage, stating that such structures “may have been present” in ancestral hominids. We are grateful to the reviewer for this comment.

"given that isochrony universally governs music and that recursion is a feature of music, findings (sic.) suggest a possible evolutionary link between great ape loud calls and vocal music". Isochrony is also a feature of the noise produced by cicadas. Does this suggest an evolutionary link between vocal music and the noise of cicadas?

We apologise, but it is unclear what the reviewer is exactly suggesting or proposing here. It seems as though it is believed that cicadas are as phylogenetically related to humans as great apes are. Our last common ancestor with great apes diverged about 10mya, but with cicadas 600mya. The last common ancestor with great apes was a great ape (or hominid). The human-cicada last common ancestor would have looked like a worm (it is probable it would already have a nervous bulge at the head, or “brain”). In order to avoid similar misinterpretations, we have now clarified in several instances that our study and interpretation of results are based on shared ancestry within the Hominid family.

It seems that the reviewer may be also misinterpreting our findings. We do not simply report isochrony in a wild great ape (multiple references for isochronous calls in primate are provided in the Discussion). We report isochrony within isochrony in three non-exclusive rhythmic arrangements. In case the reviewer knows of a study on cicadas, or any non-human species, showing recursive sound combinatorics of this nature, we kindly request the citation. We can only hope that such new cases may be gradually unveiled in wild animals to help propel our general understanding of possible ways of how insipient recursive vocal combinatorics in ancient hominids could have given rise to recursion as used today by language-able modern humans.

Finally, some passages also reveal quite glaring misunderstandings of the cited literature. For instance:

"Therefore, the search for recursion can be made in the absence of meaning-base operations, such as Merge, and more generally, semantics and syntax". It is precisely Chomsky's (disputable) opinion that the main operation that govern syntax, Merge, has nothing to do with semantics. The latter is dealt within a putative conceptual-intentional performance system (in Chomsky's terminology), which is governed by different operations.

Following the reviewer’s comment, we have now removed “meaning-base operations, such as Merge, and more generally” from the target sentence in order to avoid confusion. Thank you.

"Namely, experimental stimuli have consisted of artificial recursive signal sequences organized along a single temporal scale (though not structurally linear), similarly with how Merge and syntax operate". The minimalist view advocated by Chomsky assumes that mapping a hierarchichal structure to a linear order (a process called linearizarion) is part of the articulatory-perceptual system. This system is likewise not governed by Merge and is not part of "syntax" as conceived by the Chomskyan minimalists.

Following the reviewer’s comment, we have not omitted the target sentence for added clarity.

Reviewer #1 (Recommendations For The Authors):

L55-67: I feel there is a step missing in the logic of the argumentation here. The studies cited by the authors here are mostly about syntactic-like structuring but not recursion. Hence when the authors mention in the next sentence that these studies investigate the perception of recursive signalling, it seems incorrect. I agree with the logic, but the references do not seem appropriate. I would further suggest that if there are no other references, that would make the introduction of the study here even easier: there is very little work investigating this capacity in non-human animals, let alone on a production perspective, therefore, the study conducted here is paramount and fills this important gap in the literature.

We are grateful to Reviewer #1 for these comments, and we are honoured to hear that our findings are filling a literature gap. We have now carefully revised the manuscript, hopefully, streamlining our line of reasoning and improving the paper’s overall readability. We agree that there is very little work investigating the spontaneous “production” of recursion in nonhuman animals. We decided to better detail the logic of our paper by clarifying the difference between recursion and repetition and clarifying that the motifs that we identify in wild orangutan represent a case of "temporal recursion".

L59: Johan J should be removed (same in discussion).

Removed, thanks.

L60: For example is repeated twice, here and L55.

We have rephrased this part of the manuscript, thanks.

L72-73: If we consider the Watson et al., 2020 study an example of recursive perception (which I do not think is true), this was conducted using a passive design - i.e. with no active training.

We have rephrased this part of the manuscript, thanks.

L240-241: Again, non-adjacent dependency processing does not equal recursion.

We agree that non-adjacent dependency processing does not equal recursion. We have now clarified this section accordingly.

L269: one of the most.

Corrected, thanks.

L296: add space after settings.

Corrected, thanks.

Reviewer #2 (Recommendations For The Authors):

In addition to the public portion of the review, I advise the authors' to substantially alter their style of writing. The language used is not accurate and the intended meaning is often not clear. This makes it hard for any reader to follow the authors' reasoning fully. Below I list only a few of the egregious examples but the examples abound:

"this hints at a neuro-cognitive or neuro-computational transformation in the human brain" what meaning do the author assign to "neuro-cognitive" and "neuro-computational" ? what difference do they place between the two (so that they would be disjoined.) ? What "transformation" are we talking about ? From what to what ?

" However, recursive signal structures can also unfold in other manners, such as across nested temporal scales and in the absence of semantics (Fitch, 2017a), as in music." what is meant here by nested temporal scales ?

"The simultaneous occurrence of non-exclusive recursive patterns excludes the likelihood that orangutans concatenate long calls and their subunits in linear structure without any recursive processes": isn't there a more straightforward way to say "excludes the likelihood"? What is meant by "non-exclusive recursive patterns"?

It seems that Reviewer #2 does not share our writing style. Nonetheless, we have tried to meet the reviewer halfway, clarifying throughout the new revised version our definitions, our line of argument, our motivations, our results, the context of our findings in what is known about recursion in animals, and the implication of our discovery for language evolution theory.

Version 2

Abstract

Abstract

Recursive procedures that allow placing a vocal signal inside another of similar kind provide a neuro-computational blueprint for syntax and phonology in spoken language and human song. There are, however, no known vocal sequences among nonhuman primates arranged in self-embedded patterns that evince vocal recursion or potential insipient or evolutionary transitional forms thereof, suggesting a neuro-cognitive transformation exclusive to humans. Here, we uncover that wild flanged male orangutan long calls feature rhythmically isochronous call sequences nested within isochronous call sequences, consistent with two hierarchical strata. Remarkably, three temporally and acoustically distinct call rhythms in the lower stratum were not related to the overarching rhythm at the higher stratum by any low multiples, which suggests that these recursive structures were neither the result of parallel non-hierarchical procedures or anatomical artifacts of bodily constrains or resonances. Findings represent a case of temporally recursive hominid vocal combinatorics in the absence syntax, semantics, phonology or music. Second-order combinatorics, ‘sequences within sequences’, involving hierarchically organized and cyclically structured vocal sounds in ancient hominids may have preluded the evolution of recursion in modern language-able humans.

Introduction

Among the many definitions of recursion (Martins, 2012), the view that it represents the repetition of an element or pattern within a self-similar element or pattern has crossed centuries and disciplines, from von Humboldt (1836) and Hockett (1960), to Mandelbrot (1980) and Chomsky (2010); from fractals in mathematics (Mandelbrot, 1980) to generative grammars in linguistics (Chomsky, 2010), from graphic (e.g., “Print Gallery” by M. C. Escher) to popular art (e.g., 1940’s Batman #8 comic book cover). Across varying terminologies, the common denominator across fields is that to re-curse (from the Latin to ‘re-run’ or ‘re-invoke’) is an operation that produces multiple, potentially infinite sets of items from one initial item or a finite set. This is achieved by nesting an item within itself or within another item of the same kind. Recursive patterns in everyday life are ubiquitous and include, for example, computer folders stored inside other computer folders, Russian dolls nested in each other, Romanesco broccoli’s spirals arranged in a spiral, and the same number of minutes passed within the same number of hours (e.g., 12:12). Accordingly, recursion is not the simple repetition of a pattern or item on single level (e.g., computer folders or Russian dolls side by side), but the placement of a pattern or item within itself (e.g., computer folders or Russian dolls inside each other), hence, generating different hierarchical levels or strata. This means that the same pattern or item is encountered at least at two different scales (e.g., 12 at the scale of hours, and 12 at the scale of minutes).

In language, although classically associated with syntax (Chomsky, 2010; Idsardi et al., 2018), recursion and its diagnostic self-embedded patterns have been recognised in phonology (Bennett, 2018; Elfner, 2015; Kabak and Revithiadou, 2009; Nasukawa, 2015, 2020; Vogel, 2012), verbal and non-verbal music (Jackendoff, 2009; Koelsch et al., 2013; Martins et al., 2017; Sharma and Chimalakonda, 2018), making these systems open-ended and theoretically inexhaustible. Recursive vocal sequences or structures in nonhuman primates could potentially inform insipient or transitional states of recursion along human evolution before the rise of modern language. Their apparent absence, notably in great apes – our closest living relatives – has been interpreted as indicating that a neuro-cognitive or neuro-computational transformation occurred in our lineage but none other (Hauser et al., 2002). This absence of evidence has led some scholars to question altogether the role of natural selection for the emergence of language, tacitly favouring sudden “hopeful monster” mutant scenarios (Berwick and Chomsky, 2019; Bolhuis and Wynne, 2009).

Decades-long debates on the evolution of language have carved around the successes and limitations of empirical comparative animal research (Bolhuis et al., 2018; Bowling and Fitch, 2015; Corballis and Corballis, 2014; Lameira, 2017; Lameira and Call, 2020; Martins and Boeckx, 2019; Rawski et al., 2021; Townsend et al., 2018). Syntax-like vocal combinatorics have been identified in some bird (Engesser et al., 2019, 2016; Suzuki et al., 2016, 2017) and primate species (Jiang et al., 2018; Wang et al., 2015; Watson et al., 2020), but vocal combinatorics were not claimed to be recursive nor was recursion directly tested. Three notable exceptions demonstrated recursion learning in non-human animals settings: Gentner et al. (2006) in European starlings, Ferrigno et al. (2020) in rhesus macaques, and Liao et al. (2022) in crows. These studies show that animals can learn to recognise recursion in synthetic stimuli after dedicated human training in laboratory settings, but they do not show spontaneous production of recursive vocal combinatorics in naturalistic settings. Evidence of recursive vocal structures in wild animals (i.e., without human priming or intervention), notably in primates closely related phylogenetic to humans, such as great apes, would better inform what evolutionary precursors and processes could have led to emergence of recursion in the human lineage.

Direct structural approach to recursive combinatorics

A novel, direct approach to recursive vocal combinatorics with wild primates is desirable to help infer signal patterns that were recursive in some degree or kind in an extinct past, and moulded subsequently into the recursive structures observed today in humans. By virtue of their own primitive nature, proto-recursive structures did not likely fall within modern-day classifications. Therefore, they will often fail to be predicted based on assumptions guided by modern language (Kershenbaum et al., 2014; Miyagawa, 2021). To this end, a structural approach is particularly advantageous based on the cross-disciplinary definition of recursion as “the nesting of an element or pattern within a self-similar element or pattern”. First, no prior assumptions are required about species’ cognitive capacities. High-level neuro-motor procedures are inferred only to the extent these are directly reflected in how signal sequences are organised. For example, Chomsky’s definition of recursion (Chomsky, 2010) can generate non-self-embedded signal structures, but these would be for that same reason operationally undetectable amongst other signal combinations. Second, no prior assumptions are required about signal meaning. There are no certain parallels with semantic content and word meaning in animals, but analyses of signal patterning allow to identify similarities between non-semantic (nonhuman) and semantic (human) combinatoric systems (Lipkind et al., 2013; Sainburg et al., 2019). The search for recursion can, hence, be made in the absence of lexical items, semantics, or syntax. Third, no prior assumptions are required about signal function. Under any evolutionary scenario, including punctuated hypotheses, ancestral signal function (whether cooperative, competitive, or otherwise) is expected to have derived or been leveraged by its proto-recursive structure. Otherwise, once present, recursion would not have been fixated among human ancestral populations. Accordingly, a structural approach opens the field to untapped signal diversity in nature and yet unrecognised bona fide combinatoric possibilities within the human clade.

Exploring recursive combinatorics in a wild great ape

Here, we undertake an explorative but direct structural approach to recursion. We provide evidence for recursive self-embedded vocal patterning in a (nonhuman) great ape, namely, in the long calls of flanged orangutan males in the wild. We conducted precise rhythm analyses (De Gregorio et al., 2021; Roeske et al., 2020) of 66 long call audio recordings produced by 10 orangutans (Pongo pygmaeus wurmbii) across approximately 2510 observation hours at Tuanan, Central Kalimantan, Indonesian Borneo. We identified 5 different element types that comprise the structural building blocks of long calls in the wild (Hardus et al., 2009; Lameira and Wich, 2008), of which the primary type are full pulses (Fig. 1A). Full pulses do not, however, always exhibit uninterrupted vocal production throughout a long call [as during a long call’s climax (Spillmann et al., 2010)] but can break-up into 4 different “sub-pulse” element types: (i) grumble sub-pulses [quick succession of staccato calls that typically constitute the first build-up pulses of long calls (Hardus et al., 2009)], (ii) sub-pulse transitory elements and (iii) pulse bodies (typically constituting pulses before and/or after climax pulses) and (iv) bubble sub-pulses (quick succession of staccato calls that typically constitute the last tail-off pulses of long calls) (Fig. 1A). We characterised long calls’ full- and sub-pulses’ rhythmicity to determine if orangutan long calls present a re-iterated structure across different hierarchical strata. We extracted inter-onset-intervals (IOIs; i.e., time difference between the start of a vocal element and the preceding one - tk) from 8993 vocal long call elements (Fig. 1A): 1930 full pulses (1916 after filtering for 0.025<tk<5s), 757 grumble sub-pulses (731), 1068 sub-pulse transitory elements (374), 816 pulse bodies (11) and 4422 bubble sub-pulses (4193). From the extracted IOIs, we calculated their rhythmic ratio by dividing each IOI by its duration plus the duration of the following interval. We then computed the distribution of these ratios to ascertain whether the rhythm of long call full and sub-pulses presented natural categories, following published protocols (De Gregorio et al., 2021; Roeske et al., 2020) (Fig. 1B, C, D).

Organization and rhythmic features of orangutans’ long calls.

(A) On top, the spectrogram of a Full Pulse and its organization in Sub-Pulses (e.g., Grumble sub-pulses). Below are the spectrograms of the three other sub-element types: Sub-pulse transitory elements, Pulse bodies and Bubble sub-pulses. Bars on the top of each spectrogram schematically quantify durations of inter-onset intervals (tk): dark green denotes the higher-level of organization (Full pulse). Orange (in the inset) and light green (bottom right) denote the lower-level organization (sub-pulse element types).

(B) Probability density function showing the distributions of the inter-onset-intervals (tk) for each of the long call element types.

(C) The distributions on the left show rhythm ratios (rk) per element type as calculated on 12 flanged males for a total of 1915 Full-pulses and 5309 sub-pulses. Solid sections of the curves indicate on-isochrony rk values; striped sections indicate off-isochrony rk values. A solid white line indicates the 0.5 rk value corresponding to isochrony. White dotted lines denote the on-isochrony peak value extracted from the probability density function. On the right, a bar plot per each element type shows the percentage of observations (rk) falling into the on-isochrony boundaries (solid bars) or on off-isochrony boundaries (striped bars). The number of on-isochrony rk is significantly larger (GLMM, Full vs Null: Chisq=2717.543, p<0.001) than the number of off-isochrony rk for all long call element types (Full pulse: t-ratio=-25.164, p<0.001; Bubble sub-pulse: t-ratio= -30.694, p<0.001; Grumble sub-pulse: t-ratio=-14.526, p<0.001; Sub-pulse transitory element: t-ratio=-3.148, p<0.001). Pulse body showed no rk values falling within the on-off-isochrony boundaries.

(D) Distribution of a variable calculated as the ratio between the tk of a sub-pulse and the tk of the corresponding higher level of organization, the Full Pulse. We report the peak value of the curve (0.046) and tested the significance of the extent of the central quartiles, which was significantly smaller than peripheral quartiles (Wilcoxon signed-rank test: W=2272, p<0.001).

Results

The density probability function of orangutan full pulses showed one peak (rk=0.493) in close vicinity to a theoretically pure isochronic rhythm, that is, full pulses were regularly paced at 1:1 ratio, following a constant tempo along the long call (Fig. 1C). Our model (GLMM, full model vs null model: Chisq=298.2876, df=7, p<0.001; see Supplementary Materials) showed that pulse type, range of the curve (on-off-isochrony), and their interaction, had a significant effect on the count of rk values. In particular, full pulses’ isochronous peak tested significant (t.ratio=-15.957, p<0.0001), that is, the number of rk values falling inside on-isochrony range was significantly higher than the number of rks falling inside the off-isochrony range (Fig. 1C). Critically, three (of the four) orangutan sub-pulse element types – grumble sub-pulses, sub-pulse transitory elements and bubble sub-pulses – also showed significant peaks (grumble sub-pulses: t.ratio = -5.940, p<.0001; sub-pulse transitory elements: t.ratio=-4.048, p=0.0001; bubble sub-pulses: t.ratio= - 10.640, p<.0001) around pure isochrony (peak rk: grumble sub-pulses = 0.501; sub-pulse transitory elements=0.495; bubble sub-pulses=0.502; Fig. 1C). That is, sub-pulses were regularly paced within regularly paced full pulses, denoting isochrony within isochrony (Fig. 1C) at different average tempi (mean tk (sd): full pulses=1.696 (0.508); grumble sub-pulses=0.118 (0.111); sub-pulse transitory elements=0.239 (0.468); bubble sub-pulses= 0.186 (0.292); Fig. 1B). Overall, sub-pulses’ tk was equivalent to 0.046 of their comprising full-pulses (Fig. 1D), which put sub-pulses at an approximate ratio of 1:22 relative to that of full-pulses, the smallest categorical temporal rhythmic interval registered thus far in a vertebrate (De Gregorio et al., 2021; Roeske et al., 2020).

Permuted discriminant function analyses (Mundry and Sommer, 2007) (crossed, in order to control for individual variation) in R (Team, 2013) based on seven acoustic measures extracted from grumble, transitory elements, and bubble sub-pulses confirmed that these represented indeed acoustically distinct sub-pulse categories, where the percentage of correctly classified selected cases (62.7%) was significantly higher (p=0.001) than expected (37%).

Discussion

Rhythmic analyses of orangutan long calls reveal the presence of self-embedded isochrony in the vocal combinatorics of a wild great ape. Notably, we found that wild orangutan long calls exhibit two discernible structural strata – the full- and sub-pulse level – and three non-exclusive nested motifs in the form of [isochronyA [isochronya,b,c]](Fig. 2).

Isochrony nested within isochrony.

Three acoustically distinct sub-pulse calls occurring at three distinct tempi nested within the same pulse-level tempo in wild flanged male orangutan long calls.

This is fundamentally distinct from a simple repetition of calls or call isochrony – when a call repeats linearly at a constant interval – which are common features in some animal sound communication systems (De Gregorio et al., 2023). Instead, we demonstrate how a vocal element repeated at a constant interval is itself composed by (one of three possible) vocal elements that also repeat themselves at a constant interval of different tempi.

The orangutans’ production of recursive vocal motifs in the wild, and therefore, without training, is especially compelling in the context of the lab-based work that shows that nonhuman animals can learn recursion with training (Ferrigno et al., 2020; Gentner et al., 2006; Liao et al., 2022). Some aspects of these vocal combinatoric structures could be potentially learned as well (Lameira et al., 2022, 2016, 2015; Lameira and Shumaker, 2019; Wich et al., 2012), but this study is agnostic on this matter because its design does not allow to single out learning effects. Nonetheless, results show that temporal recursion occurs spontaneously in the wild in great ape vocal communication.

Can great apes hear recursive isochrony?

The observation that the long calls of orangutans in nature possess isochronous characteristics raises questions about the ability of apes to perceive isochronous signals. Humans perceive an acoustic pulse as a continuous pitch, instead of a rhythm, at rates higher than 30 Hz (i.e., 30 beats per second). Human and nonhuman great apes have similar auditory capacities (Quam et al., 2015) and there are limited skeletal differences in inner ear anatomy that suggest significantly distinct sensitivity, resolution, or activation thresholds in the time domain (Quam et al., 2015; Spoor and Zonneveld, 1998). Long call sub-pulses exhibited average rhythms at ∼9.263 (sd: 3.994) Hz [i.e., tk=0.184 (0.303) s]. Therefore, ear anatomy offers confidence that orangutans (and other great apes), like humans, perceive sub-pulse rhythmic motifs at these rates as such, that is, a train of signals, instead of one uninterrupted signal. Assuming otherwise would imply that auditory time-resolution differs by more than one order of magnitude between humans and other great apes in the absence of obvious anatomical culprits.

Can physiology fully explain recursive isochrony?

The occurrence of three non-exclusive recursive patterns (i.e., three acoustically distinct sub-pulse calls occurring at three distinct tempi nested within the same pulse-level tempo), substantially decreases the probability that recursion was the primary by-product of anatomic constrains, such as vocal fold oscillation, breath length, heartbeat, and other physiological processes or movements (Pouw et al., 2020). Such processes can generate frequency patterns nested within others, however, in these cases sub-frequencies occur in the form of harmonics related to the reference (dominant) frequency and to each other by small whole-numbered multiples. Yet, the three observed rhythmic arrangements at the sub-pulse level were not related to the pulse level by any small integer ratios (i.e., 1/22). Also, some of these processes (e.g., vocal fold action) are oscillatory in nature, involving nested frequency waves. They are not combinatorial, involving nested sequences of events, as we report here.

Our data stimulate new questions about the relationship between oscillators and combinatoriality, which is difficult to investigate from an observational point view in the wild, but our results will hopefully inspire new studies using controlled experimental settings to assess how oscillators and combinatoriality may be associated in ways potentially richer than thus far suspected. Together, our findings suggest that recursive isochrony is not the absolute result of raw mechanics but is instead likely generated or tampered with by, at least, a temporally recursive neuro-motor procedure.

Can a linear algorithm produce recursive isochrony?

The occurrence of three non-exclusive recursive patterns drives down the likelihood that orangutans concatenate long call pulses and sub-pulses in linear fashion and without bringing into play a recursive neuro-motoric process. To generate the observed vocal motifs linearly, three independent neuro-computational procedures would need to run in parallel. These three independent procedures would need to be indistinguishable, transposable, and/or interchangeable at the pulse level, whilst generating distinct isochronic rhythms and acoustics at the sub-pulse level. If theoretically possible at all, one would predict some degree of interference between the three linear procedures at the pulse level, manifested in some of form of deviation around the isochrony peak. However, this was not observed; distribution of data points on and off isochrony was equivalent between pulses and sub-pulses.

Precursor forms are not modern forms

Recursive self-embedded vocal motifs in orangutans indicate that vocal recursion among hominids is not exclusive to human vocal combinatorics, at least in the form of temporally embedded regular rhythms. This is not to suggest that orangutan recursive motifs exhibit all other properties that recursion exhibits in modern language-able humans, or that the two are the same, or equivalent. Further research will be necessary to fully unveil how orangutans use and control vocal recursion to form a clearer evolutionary picture. Expecting equivalence with language is, however, unwarranted as it would imply that no evolution has occurred in over 10 million years since the split between orangutans and humans. Any differences between our findings and recursion in today’s syntax, phonology, or music do not logically reject the possibility that recursive isochrony represents an ancient, or perhaps ancestral, state for the evolution of vocal recursion within the Hominid family.

Implications for the evolution of recursion

Recursion and fractal phenomena are prevalent across the universe. Celestial and planetary movement, the splitting of tree branches, river deltas and arteries, the morphology of bacteria colonies. Patterns within self-similar patterns are the norm, not the exception. This makes the seeming singularity of human recursion amongst animal vocal combinatorics all the more enigmatic. The discovery of recursive vocal patterns organized along two hierarchical temporal levels in a hominid besides humans suggests that ‘sequences within sequences’ may have been present in ancestral hominids, and hence, that they may have predated the emergence of language in the human lineage.

Three major implications for the evolution of recursion in language apply. First, much ink has been laid on the topic. Yet, the possibility of self-embedded isochrony, or non-exclusive self-embedded patterns occurring within the same signal sequence, has on no account been formulated or conjectured as a possible state of recursive signalling, be it in vertebrates, mammals, primates, or otherwise, extant or extinct. This suggests that controversy may have been underscored by data-poor circumstances on vocal combinatorics in wild great apes, which only now start gathering comprehensive research effort (Bortolato et al., 2023b, 2023a; Girard-Buttoz et al., 2022; but see Lameira et al., 2013a). Resolution may come through a re-evaluation of previous studies with further related taxa and with experimental tests designed within a richer and more articulated panorama of observations on vocal combinatorics in wild great apes. Recursive vocal patterning in a wild great ape in the absence of syntax, semantics, phonology, or music opens a new charter for possible insipient and transitional states of recursion among hominids. The open discussion of what properties make a structure proto-recursive will be essential to move the state-of-knowledge past antithetical, dichotomous notions of how recursion and syntax evolved (Berwick and Chomsky, 2019; Martins and Boeckx, 2019).

Second, our findings invite renewed interest and re-analysis of primate vocal combinatorics in the wild (Gabrić, 2021; Girard-Buttoz et al., 2022). Given the dearth of such data, findings imply that it may be too hasty to discuss whether combinatorial capacities in primates or birds are equivalent to those engaged in syntax (Engesser et al., 2015; Watson et al., 2020) or phonology (Bowling and Fitch, 2015; Rawski et al., 2021). Such classifications may be putting the proverbial cart before the horse; they are based on untested assumptions that may not have applied to proto-recursive ancestors (Kershenbaum et al., 2014; Miyagawa, 2021), for example, that syntax and phonology evolved as separate “modules”, that one attained modern form before the other, or that they evolved in hominids regardless of whether consonant-like and vowel-like calls were present or not.

Third, given that isochrony universally governs music and that recursion is a feature of music, findings could suggest a possible evolutionary link between great ape loud calls and vocal music. Loud calling is an archetypal trait in primates (Wich and Nunn, 2002). Our findings suggest that among ancient hominids, loud calling may have preceded, and subsequently transmuted, into modern recursive vocal structures in humans found today in the form of song or chants. Given their conspicuousness, loud calls represent one of the most studied aspects of primate vocal behaviour (Wich and Nunn, 2002), but their rhythmic patterns have only recently started to been characterized with precision (Clink et al., 2020; De Gregorio et al., 2021; Gamba et al., 2016). Besides our analyses, there are remarkably few confirmed cases of vocal isochrony in great apes (but see Raimondi et al., 2023), but the behaviours that have been rhythmically measured with accuracy have been implicated in the evolution of percussion (Fuhrmann et al., 2015) and musical expression (Dufour et al., 2015; Hattori and Tomonaga, 2020), such as social entrainment in chimpanzees in connection with the origin of dance (Lameira et al., 2019) [a capacity once also assumed to be neurologically impossible in great apes (Fitch, 2017; Patel, 2014)]. This opens the intriguing, tentative possibility that recursive vocal combinatorics were first and foremost a feature of proto-musical expression in human ancestors, later recruited and “re-engineered” for the generation of linguistic combinatorics.

Concluding remarks

The presence of temporally recursive vocal motifs in a wild great ape revolutionizes how we can approach the evolution of recursion along the human lineage beyond all-or-nothing accounts. Future studies on primate vocal combinatorics, particularly undertaking a structural approach and in the wild, offer promising new paths to empirically assess possible precursors and proto-states for the evolution of recursion within the Hominid family, also adding temporal recursion as a new layer of analysis. These crucial data on the evolution of recursion, language, and cognition along the human lineage will materialise if, as stewards of our planetary co-habitants, humankind secures the survival of nonhuman primates and the preservation of their habitats in the wild (Estrada et al., 2022, 2017; Laurance, 2013; Laurance et al., 2012).

Methods and Materials

Study site

We conducted our research at the Tuanan Research Station (2°09′S; 114°26′E), Central Kalimantan, Indonesia. Long calls were opportunistically recorded from identified flanged males (Pongo pygmaeus wurmbii) using a Marantz Analogue Recorder PMD222 in combination with a Sennheiser Microphone ME 64 or a Sony Digital Recorder TCD-D100 in combination with a Sony Microphone ECM-M907.

Acoustic data extraction

Audio recordings were transferred to a computer with a sampling rate of 44.1 kHz. Seven acoustic measures were extracted directly from the spectrogram window (window type: Hann; 3 dB filter bandwidth: 124 Hz; grid frequency resolution: 2.69 Hz; grid time resolution: 256 samples) by manually drawing a selection encompassing the complete long call (sub)pulse from onset to offset, using Raven interactive sound analysis software (version 1.5, Cornell Lab of Ornithology). These parameters were duration(s), peak frequency (Hz), peak time, peak frequency contour average slope (Hz), peak frequency contour maximum slope (Hz), average entropy (Hz), signal-to-noise ratio (NIST quick method). Please see software’s documentation for full description of parameters (https://ravensoundsoftware.com/knowledge-base/pitch-tracking-frequency-contour-measurements/). Acoustic data extraction complemented the classification of long calls elements, both at the pulse and sub-pulse levels, based on close visual and auditory inspection of spectrograms, both based on elements’ distinctiveness between each other as well as in relation to the remaining catalogued orangutan call repertoire (Hardus et al., 2009) (see also supplementary audio files). Of these parameters, duration and peak frequency in particular have been shown to be resilient across recording settings(Lameira et al., 2013b) and to adequately represent variation in the time and frequency axes (Lameira et al., 2017).

Rhythm data analyses

Inter-onset-intervals (IOI’s = tk) were only calculated from the begin time (s) of each full- and sub-pulse long call elements using Raven interactive sound analysis software, as above explained. tk was calculated only from subsequent (full/sub) pulse elements of the same type. Ratio values (rk) were calculated as tk/(tk+tk+1). Following the methodology of Roeske et al., 2020 and De Gregorio et al. 2021, to assess the significance of the peaks around isochrony (corresponding to the 0.5 rk value), we counted the number of rks falling inside on-isochrony ranges (0.440 < rk < 0.555) and off-isochrony ranges (0.400 < rk < 0.440 and 0.555 < rk < 0.600), symmetrically falling at the right and left sides of 1:1 ratios (0.5 rk value). We tested the count of on-isochrony rks versus the count of off-isochrony rks, per pulse type, with a GLMM for negative-binomial family distributions, using glmmTMB R library. In particular, we built a full model with the count of rk values as the response variable, the pulse type in interaction with the range the observation fell in (on-or off-isochrony) as predictors. We added an offset weighting the rk count based on the width of the bin. The individual contribution was set as random factor. We built a null model comprising only the offset and the random intercepts. We checked the number of residuals of the full and null models, and compared the two models with a likelihood ratio test (Anova with “Chisq” argument).

We calculated p-values for each predictor using the R summary function and performed pairwise comparisons for each level of the explanatory variables with emmeans R package, adjusting all p-values with Bonferroni correction. We checked normality, homogeneity (via function provided by R. Mundry), and number of the residuals. We checked for overdispersion with performance R package (Lüdecke et al., 2020). Graphic visualization was prepared using R (Team, 2013) packages ggplot2 (Wickham, 2009) and ggridges (Wilke, 2022). Data reshape and organization were managed with dplyr and tidyr R packages.

Acoustic data analyses

Permutated discriminant function analysis with cross classification was performed using R and a function provided by Roger Mundry (Mundry and Sommer, 2007). The script was: pdfa.res=pDFA.crossed (test.fac=“Sub-pulse-type”, contr.fac=“Individual.ID”, variables=c(“Delta.Time”, “Peak.Freq”, “Peak.Time”, “PFC.Avg.Slope”, “PFC.Max.Slope”, “Avg.Entropy”, “SNR.NIST.Quick”), n.to.sel=NULL, n.sel=100, n.perm=1000, pdfa.data=xdata). These analyses assured that long call elements, at the pulse and sub-pulse level, indeed represented biologically distinct categories.